Using Incremental Many-to-One Queries to Build a Fast and Tight Heuristic for A* in Road Networks

### 2021-11-11 paper/ch_potentials.tex
A* text changes
wordsmithing on bidir A*

On road graphs, bidirectional search provides an easy way to halve the running time of Dijkstra's algorithm.
A bidirectional variant of A* therefore also seems desirable.
However, as shown in~\cite{gh-cspas-05}, the necessary modifications are not straight forward.
We revisit bidirectional A* and propose an alternative approach.
Our experiments show that it is competitive with the solution described by~\cite{gh-cspas-05}.

The straightforward idea is to use two heuristics $\overrightarrow{h}_t(x)$ and $\overleftarrow{h}_s(x)$.
The forward search has its queue ordered by $\overrightarrow{D}[x] + \overrightarrow{h}_t(x)$, where $\overrightarrow{h}_t(x)$ estimates the distance $\dist(x,t)$ from $x$ to $t$.
Similarly, the backward search has its queue ordered by $\overleftarrow{D}[x] + \overleftarrow{h}_s(x)$, where $\overleftarrow{h}_s(x)$ is an estimate of the distance $\dist(s,x)$ from $s$ to $x$.
% We denote by $\overrightarrow{D}[x]$ the tentative distance of the forward search at $x$ and similarly, by $\overleftarrow{D}[x]$ the tentative distance in the backward search.
We denote by $\overrightarrow{k}$ and $\overleftarrow{k}$ the minimum queue keys and by $\mu$ the length of the shortest $st$-path found so far.

The problem with this straightforward approach is that these two potentials induce different reduced graphs (see Section~\ref{sec:a_star}).
Thus, in the equivalent bidirectional Dijkstra search each direction would run on a different graph.
This breaks the usual bidirectional Dijkstra stopping criterion.
To the best of our knowledge, no better stopping criterion exists than running \emph{both} searches until the unidirectional stopping criterion is met for one direction.
Thus, this straightforward bidirectional A* performs roughly twice the work of a unidirectional A*.
This is not a useful algorithm.
However, it can serve as basis for further algorithmic refinements.
The authors of~\cite{gh-cspas-05} therefore refer to this approach as \emph{symmetric} bidirectional A*.

To obtain a bidirectional stopping criterion, the \emph{average potential} is proposed~\cite{gh-cspas-05}.
It combines a forward and a backward heuristic $\overrightarrow{h}_t$ and $\overleftarrow{h}_s$ into a combined \emph{average heuristic} $h_{\overline{st}}$.
The idea is to use a common reduced graph, whose weights are the average weights of the individual reduced graphs.
In this case, both searches run on the same graph.
This allows stopping the searches when $\overrightarrow{k} + \overleftarrow{k} \geq \mu$.
Formally, $h_{\overline{st}}(x)$ is defined as $(\overrightarrow{h}_t(x) - \overleftarrow{h}_s(x))/2$.
The forward search uses $h_{\overline{st}}(x)$ as heuristic.
The backward search uses $-h_{\overline{st}}(x)$.
Unfortunately, average potentials have two downsides.
First, evaluating the average potential requires evaluating both $\overrightarrow{h}_t$ and $\overleftarrow{h}_s$.
Evaluating the average heuristic is therefore slower than evaluating just one the composing heuristics.
Second, the bidirectional stopping criterion comes at the cost of worse estimates for each direction on its own.
$h_{\overline{st}}(x)$ is a worse estimate for $\dist(x,t)$ than $\overrightarrow{h}_t$.
Similarly, $-h_{\overline{st}}(x)$ is a worse estimate for $\dist(s,x)$ than $\overleftarrow{h}_s$.
The second downside can be partially mitigated through pruning with the composing heuristics.
When the forward search scans an edge $xy$ where $\overrightarrow{D}[x] + w(x,y) + \overrightarrow{h}_s(y) > \mu$ holds, i.e. the distance plus the estimate of the remaining distance is already greater than the currently known tentative distance, $y$ needs not to be inserted into the queue.\todo{does not have to be inserted? needs not klingt komisch... aber English ist nicht meine Stärke. :) }
The pruning rule for the backward search is analogous.

\todo{$w(xy)$ vs $w(x,y)$}
To avoid the downsides of the average potential, we revisit symmetric bidirectional A* and propose a new pruning criterion.
We describe the idea for the forward search.
The pruning rule for the backward search is analogous.
The central idea consists of using information from the backward search to prune edges in the forward search.
We do not use the average heuristic.
Instead of a strong stopping criteria, we use a pruning rule that gets stronger the longer the search runs.
Such a pruning rule will eventually prune all remaining branches and stop the search.
Formally, we use the undirectional stopping criteria.

Let $xy$ be an edge that we relax in the forward search.
Before pushing $y$ into the queue, we apply the pruning rule.
If we can prove that every path using $xy$ is at least as long as the shortest known path length $\mu$, then we do not have to push $y$.
We therefore want to obtain a lower bound for $\dist(s,x)+w(x,y)+\dist(y,t)$.
As $x$ was settled, $\overrightarrow{D}[x]$ contains the shortest path length $\dist(s,x)$, i.e., $\dist(s,x)=\overrightarrow{D}[x]$.
$w(x,y)$ is also exactly known as it's just an edge weight.
It remains to lower bound $\dist(y,t)$.
Nodes are removed from the backward queue ordered by $\overrightarrow{h}(y) + \dist(y,t)$.
\todo{$\ge$ vs $\geq$ und $\le$ vs $\leq$, macht das einen Unterschied?}
If $y$ was not yet removed from the backward queue, we know that $\overrightarrow{h}(y) + \dist(y,t) \geq \overleftarrow{k}$.
This gives us the required lower bound $\dist(y,t) \geq \overleftarrow{k}-\overrightarrow{h}(y)$.
$y$ does not have to be pushed, if it was not settled by the backward search and $\overrightarrow{D}[x]+w(x,y)+\overleftarrow{k}-\overrightarrow{h}(y)\ge \mu$.
This does not mean that $y$ will never be pushed into the forward queue.
It is possible that another edge $zy$ exists for which the pruning rule cannot be applied.

\todo{Latex mag overrightarrow nicht in todos... Ich packe die Frage deswegen inline in den Text.}
Kann man sinnvoll prunen wenn $y$ von der Rückwärtssuche gesettled wurde? Z.b. Kante $xy$ prunen wenn $x$ forward gestelled ist und $y$ backward gesettled ist und $\overrightarrow{D}[x]+w(x,y)+\overleftarrow{D}[y]\ge \mu$ erfüllt ist.

\todo{Der folgende Paragraph ist die Vorgängerversion vom Paragraph davor. Wenn der neue Paragraph besser ist, dann kann dieser weg.}
We use the minimum queue key $\overleftarrow{k}$ of the backward search to obtain a tighter estimate of $\dist(x,t)$ for pruning in the forward search.
Recall that queues are ordered by $D[x] + h(x)$.
If a node $x$ was not yet settled by backward search then $\dist(x,t) + \overleftarrow{h}_s(x) \geq \overleftarrow{k}$, and equivalently $\dist(x,t) \geq \overleftarrow{k} - \overleftarrow{h}_s(x)$.
Thus, in the forward search if $\overrightarrow{D}[x] + \overleftarrow{k} - \overleftarrow{h}_s(x) > \mu$, we do not need to push $x$ into the queue.
The pruning for the backwards search works analogously.
As the pruning becomes tighter as the searches progress, it compensates for the worse stopping criterion.
Since this pruning is only used when $\mu > \infty$, for the greatest part of the search each direction only has to evaluate its own heuristic.
As our experiments show, this makes the simpler symmetric approach competitive to the average potential.

>>>>

On road graphs, bidirectional search provides an easy way to halve the running time of Dijkstra's algorithm.
A bidirectional variant of A* therefore also seems desirable.
However, as shown in~\cite{gh-cspas-05}, the necessary modifications are not straightforward.
We revisit bidirectional A* and propose an alternative approach.
Our experiments show that it is competitive with the solution described by~\cite{gh-cspas-05}.

The straightforward idea is to use two heuristics $\overrightarrow{h}_t(x)$ and $\overleftarrow{h}_s(x)$.
The forward search has its queue ordered by $\overrightarrow{D}[x] + \overrightarrow{h}_t(x)$, where $\overrightarrow{h}_t(x)$ estimates the distance $\dist(x,t)$ from $x$ to $t$.
Similarly, the backward search has its queue ordered by $\overleftarrow{D}[x] + \overleftarrow{h}_s(x)$, where $\overleftarrow{h}_s(x)$ is an estimate of the distance $\dist(s,x)$ from $s$ to $x$.
We denote by $\overrightarrow{k}$ and $\overleftarrow{k}$ the minimum queue keys and by $\mu$ the length of the shortest $st$-path found so far.

The problem with this straightforward approach is that these two potentials induce different reduced graphs (see Section~\ref{sec:a_star}).
Thus, in the equivalent bidirectional Dijkstra search each direction would run on a different graph.
This breaks the usual bidirectional Dijkstra stopping criterion.
To the best of our knowledge, no better stopping criterion exists than running \emph{both} searches until the unidirectional stopping criterion is met for one direction.
The forward search can skip nodes already settled by the backward search and vice versa.
However, this straightforward bidirectional A* still performs more work than a unidirectional A*.
Thus, on its own it is not a useful algorithm.
However, it can serve as basis for further algorithmic refinements.
The authors of~\cite{gh-cspas-05} refer to this approach as \emph{symmetric} bidirectional A*.

To obtain a bidirectional stopping criterion, the \emph{average potential} is proposed~\cite{gh-cspas-05}.
It combines a forward and a backward heuristic $\overrightarrow{h}_t$ and $\overleftarrow{h}_s$ into a combined \emph{average heuristic} $h_{\overline{st}}$.
The idea is to use a common reduced graph, whose weights are the average weights of the individual reduced graphs.
Both searches run on the same common reduced graph.
This allows stopping the searches when $\overrightarrow{k} + \overleftarrow{k} \geq \mu$.
Formally, $h_{\overline{st}}(x)$ is defined as $(\overrightarrow{h}_t(x) - \overleftarrow{h}_s(x))/2$.
The forward search uses $h_{\overline{st}}(x)$ as its heuristic.
The backward search uses $-h_{\overline{st}}(x)$.
Unfortunately, average potentials have two downsides.
First, evaluating the average potential requires evaluating both $\overrightarrow{h}_t$ and $\overleftarrow{h}_s$.
Evaluating the average heuristic is therefore slower than evaluating just one of the composing heuristics.
Second, the bidirectional stopping criterion comes at the cost of worse estimates for each direction on its own.
$h_{\overline{st}}(x)$ is a worse estimate for $\dist(x,t)$ than $\overrightarrow{h}_t$.
Similarly, $-h_{\overline{st}}(x)$ is a worse estimate for $\dist(s,x)$ than $\overleftarrow{h}_s$.
The second downside can be partially mitigated through pruning with the composing heuristics.
When the forward search scans an edge $xy$ where $\overrightarrow{D}[x] + w(x,y) + \overrightarrow{h}_s(y) > \mu$ holds, i.e. the distance plus the estimate of the remaining distance is already greater than the currently known tentative distance, it is not necessary to push $y$ into the queue.
The pruning rule for the backward search is analogous.

To avoid the downsides of the average potential, we revisit symmetric bidirectional A* and propose a new pruning criterion.
We describe the idea for the forward search.
The pruning rule for the backward search is analogous.
The central idea consists of using information from the backward search to prune edges in the forward search.
We do not use the average heuristic.
Instead of a strong stopping criteria, we use a pruning rule that gets stronger the longer the search runs.
Such a pruning rule will eventually prune all remaining branches and stop the search.
The stopping criteria for each direction remain the same (unidirectional) as before.
However, often the search stops early because the queues are empty.

Let $xy$ be an edge that we relax in the forward search.
Before pushing $y$ into the queue, we apply the new pruning rule.
If we can prove that every path using $xy$ is at least as long as the shortest known path length $\mu$, then we do not have to push $y$.
We therefore want to obtain a lower bound for $\dist(s,x)+w(xy)+\dist(y,t)$.
As $x$ was settled, $\overrightarrow{D}[x]$ contains the shortest path length $\dist(s,x)$, i.e., $\dist(s,x)=\overrightarrow{D}[x]$.
$w(xy)$ is also exactly known as it's just an edge weight.
It remains to lower bound $\dist(y,t)$.
Nodes are removed from the backward queue ordered by $\overleftarrow{h}(y) + \dist(y,t)$.
If $y$ was not yet removed from the backward queue, we know that $\overleftarrow{h}(y) + \dist(y,t) \geq \overleftarrow{k}$.
This gives us the required lower bound, i.e. $\dist(y,t) \geq \overleftarrow{k}-\overleftarrow{h}(y)$.
Thus, $y$ does not have to be pushed if $\overrightarrow{D}[x]+w(xy)+\overleftarrow{k}-\overleftarrow{h}(y)\ge \mu$.
The node $y$ might still be pushed into the queue when there is another edge $zy$ for which the pruning is not possible.
Checking the pruning rule requires evaluating the backward heuristic.
However, pruning is only possible once the searches have met, i.e. $\mu < \infty$.
Before that, each direction only has to evaluate its own heuristic.
Thus, our pruning improves on both downsides of the average potential.
